$K$-corrections and filter transformations \\ from the restframe ultraviolet to the near infrared

### 2005-07-07 docs/paper/kcorrect.tex
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@406 75dff457-8751-0410-ab1a-cd701654ce6b
add
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@411 75dff457-8751-0410-ab1a-cd701654ce6b
changes
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@450 75dff457-8751-0410-ab1a-cd701654ce6b

\shortauthors{Blanton {\it et al.} (2000)}
\shorttitle{$K$-corrections and photometric redshifts}

% TODO:
% finish final templates
% fix kphotoz banding
% fit SFHs to final spectra to get stellar masses, mean ages, 
% mean metallicities

@@

\title{$K$-corrections and photometric redshifts in SDSS and 2MASS$^1$}

\author{
Michael R. Blanton\altaffilmark{\ref{NYU}}
%Tamas Budavari\altaffilmark{\ref{JHU}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%J.~Brinkmann\altaffilmark{\ref{APO}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Mamoru Doi\altaffilmark{\ref{Tokyo}},
%Daniel Eisenstein\altaffilmark{\ref{Arizona}},
%Masataka Fukugita\altaffilmark{\ref{CosmicRay},\ref{IAS}},
%James E. Gunn\altaffilmark{\ref{Princeton}},
%David W. Hogg\altaffilmark{\ref{NYU}}, and
%David J. Schlegel\altaffilmark{\ref{Princeton}}
%Julianne Dalcanton\altaffilmark{\ref{UW}},
%Jon Loveday\altaffilmark{\ref{Sussex}},
%Michael A. Strauss\altaffilmark{\ref{Princeton}},
%Mark SubbaRao\altaffilmark{\ref{Chicago}},
%David H. Weinberg\altaffilmark{\ref{Ohio}},
%John E. Anderson, Jr.\altaffilmark{\ref{Fermilab}},
%James Annis\altaffilmark{\ref{Fermilab}},
%Neta A. Bahcall\altaffilmark{\ref{Princeton}},
%Mariangela Bernardi\altaffilmark{\ref{Chicago}},
%Robert J. Brunner\altaffilmark{\ref{Caltech}},
%Scott Burles\altaffilmark{\ref{Fermilab}},
%Larry Carey\altaffilmark{\ref{UW}},
%Francisco J. Castander\altaffilmark{\ref{Chicago}, \ref{Pyrenees}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Douglas Finkbeiner\altaffilmark{\ref{Berkeley}},
%Scott Friedman\altaffilmark{\ref{JHU}},
%Joshua A. Frieman\altaffilmark{\ref{Fermilab}},
%G. S. Hennessy\altaffilmark{\ref{USNO}},
%Robert B. Hindsley\altaffilmark{\ref{USNO}},
%Takashi Ichikawa\altaffilmark{\ref{Tokyo}},
%\v{Z}eljko Ivezi\'{c}\altaffilmark{\ref{Princeton}},
%Stephen Kent\altaffilmark{\ref{Fermilab}},
%G. R.~Knapp\altaffilmark{\ref{Princeton}},
%D. Q.~Lamb\altaffilmark{\ref{Chicago}},
%R. French Leger\altaffilmark{\ref{UW}},
%Daniel C. Long\altaffilmark{\ref{APO}},
%Robert H. Lupton\altaffilmark{\ref{Princeton}},
%Timothy A.~McKay\altaffilmark{\ref{Michigan}},
%Avery Meiksin\altaffilmark{\ref{Edinburgh}},
%Aronne Merelli\altaffilmark{\ref{Caltech}},
%Jeffrey A. Munn\altaffilmark{\ref{USNO}},
%Vijay Narayanan\altaffilmark{\ref{Princeton}},
%Matt Newcomb\altaffilmark{\ref{CarnegieMellon}},
%R. C. Nichol\altaffilmark{\ref{CarnegieMellon}},
%Sadanori Okamura\altaffilmark{\ref{Tokyo}},
%Russell Owen\altaffilmark{\ref{UW}},
%Jeffrey R.~Pier\altaffilmark{\ref{USNO}},
%Adrian Pope\altaffilmark{\ref{JHU}},
%Marc Postman\altaffilmark{\ref{STScI}},
%Thomas Quinn\altaffilmark{\ref{UW}},
%Constance M. Rockosi\altaffilmark{\ref{Chicago}},
%Donald P. Schneider\altaffilmark{\ref{PennState}}, 
%Kazuhiro Shimasaku\altaffilmark{\ref{Tokyo}},
%Walter A. Siegmund\altaffilmark{\ref{UW}},
%Stephen Smee\altaffilmark{\ref{Maryland}},
%Yehuda Snir\altaffilmark{\ref{CarnegieMellon}},
%Chris Stoughton\altaffilmark{\ref{Fermilab}},
%Christopher Stubbs\altaffilmark{\ref{UW}},
%Alexander S.~Szalay\altaffilmark{\ref{JHU}},
%Gyula P.~Szokoly\altaffilmark{\ref{Potsdam}},
%Aniruddha R.~Thakar\altaffilmark{\ref{JHU}},
%Christy Tremonti\altaffilmark{\ref{JHU}},
%Douglas L. Tucker\altaffilmark{\ref{Fermilab}},
%Alan Uomoto\altaffilmark{\ref{JHU}},
%Dan vanden Berk\altaffilmark{\ref{Fermilab}},
%Michael S. Vogeley\altaffilmark{\ref{Drexel}},
%Patrick Waddell\altaffilmark{\ref{UW}},
%Brian Yanny\altaffilmark{\ref{Fermilab}},
%Naoki Yasuda\altaffilmark{\ref{NAOJ}},
%and Donald G.~York\altaffilmark{\ref{Chicago}}
}

@@

\begin{abstract}
We present a method to develop templates for calculating
$K$-corrections and photometric redshifts suitable for ultraviolet,
optical, and infrared observations. Since we base the templates on
stellar population synthesis results, the results are intepretable in
terms of approximate stellar masses and star-formation histories. We
present templates fit with this method to the Sloan Digital Sky Survey
and Two-Micron All Sky Survey photometry.  The software that generates
these results is publicly available and easily adapted to handle a
wide range of galaxy observations.
\end{abstract}

@@

\section{Fitting SEDs to broad-band magnitudes}
\label{sedfit}

First, we describe how we reconstruct galaxy SEDs from broad band
magnitude measurements.

A broad-band flux $F_k$ in filter $k$ is a measurement of the
projection of the source spectrum $f(\lambda)$ on the response
$R_k(\lambda)$ (where typically this function includes the effects of
the atmosphere, telescope, filter $k$, and the CCD itself)
\begin{equation}
F_k = \int_0^\infty d\lambda f(\lambda) R_k(\lambda).
\end{equation}
Now, let us model $f(\lambda)$ as a sum of many template spectra:
\begin{equation}
f(\lambda) = \sum_j a_j v_j(\lambda)
\end{equation}
In order to fit the $a_j$ given the observations in the bands $k$ for
a single object, we minimize
\begin{eqnarray}
\label{chi2}
\chi^2 &=& \sum_{k} \left[ \frac{F_{k,\mathrm{obs}} - F(k)}{\sigma_k}
\right]^2 \cr
&=& \sum_{ik} \left[ \frac{F_{k,\mathrm{obs}} -\sum_j a_j
  v_j(\lambda)}{\sigma_k} \right]^2
\end{eqnarray}
over the $a_j$, where $\sigma_k$ is the standard deviation of the
error in band $k$. Since we will use $v_j(\lambda)$ which are the light
emitted from stars and ionized gas, we make the reasonable assumption
that these only {\it add} light to the spectrum, not subtract it, and
require nonnegativity: that $a_j\ge 0$ for all $j$. (We discuss dust
below).  In order to perform this nonnegative minimization of Equation
\ref{chi2}, we use the method of \citet{sha02a}.

At this point, the reader probably realizes that the number of
templates $j$ should not be much higher than the number of bands $k$,
or else the problem becomes very degenerate. In fact, for the case of
unconstrained minimization of Equation \ref{chi2} the problem is
formally degenerate if $j>k$. The nonnegativity constraint adds enough
information that this is no longer the case, but still one cannot get
strong constraints on a huge number of parameters given a small number
of measurements. Nevertheless, below we {\it will} use a huge number
of $j$ given the number of $k$, simply for the purposes of finding the
subspace of possible galaxy spectra which actually conform to galaxy
observations. We do not mean to imply that we can determine each of
the resulting parameters at high signal-to-noise.

For our model templates we use a set of spectra from the stellar
evolution synthesis results of PEGASE2 (\citealt{fioc97a}). We use 15
bursts of varying ages. The top panel of Figure \ref{sfh} shows the
star-formation history of each of these bursts. For each age we create
five templates with five different metallicities ($Z=0.03$, $0.02$,
$0.008$, $0.004$, and $0.001$). The bottom panel of Figure \ref{sfh}
shows the template for the $Z=0.008$ case at every other age, starting
with the oldest. The bottom panel of For each age and metallicity, we
create three templates with three different extinctions ($\tau_V=0$,
$1$, and $3$) using the Milky Way extinction law using the ``clumpy,''
``dusty'' model of \citet{witt00a}. This procedure results in 225
templates and spans a very large range of possible spectra. Our
procedure is designed to take this large dimensionality space and
determine a low dimensionality subspace in which galaxies reside.

We fit these models to a set of galaxies for which we know the
redshifts. The majority of the redshifts are from the SDSS, but a
small number of the redshifts are from CNOC2 ({\bf ref}). We consider
only redshifts with 1$^\circ$.25 of the J2000 Equator, between
$-60<\alpha<100$. We match the locations of these spectra to the SDSS
photometric catalog ({\bf describe}) in all of the observations in
this region and to the 2MASS Extended Source Catalog (XSC). In
passing, we note that there are large errors (many arcseconds) in the
catalog astrometry for CNOC2, and we needed to take care to account
for these errors. For spectra which had matches in multiple runs of
the SDSS, we take the (unweighted) mean of their photometry to reduce
errors {\bf how many is this done for?}. We select a random set of
these galaxies in such a way that the number per redshift is nearly
constant across redshift. For each galaxy, we consider its
extinction-corrected model photometry. Figure \ref{gmrz} shows the
observed $g-r$ colors and redshifts of the resulting set of
galaxies. Note that only for the lower redshift galaxies do we have
matches to the 2MASS XSC.

Figure \ref{kfit} shows the residuals between the observed and fit
colors after minimizing Equation \ref{chi2} for all the galaxies in
our sample. Interestingly, even this large set of parameters cannot
perfectly describe the trends of color with redshift. There are
deviations at the percent level or so in $ugriz$. The $J$-band 2MASS
observations at moderate redshift appear to be about $0.1$ magnitudes
off of the models; this discrepancy is probably related to the similar
discrepancies found by \citet{blanton03d} and \citet{baldry03a} when
comparing the luminosity density of the universe to models.  Our
attitude in this paper will be to trust the data in preference to the
models and so below we describe how we deal with these discrepancies.

Each many-parameter fit to the broad-band fluxes of each galaxy yields
a complete spectrum from which to derive $K$-corrections. It also
provides a value of $\chi^2$; in principle, for galaxies without
redshifts we can find the assumed redshift with minimum $\chi^2$ in
order to find the photometric redshift. In practice, the fit takes too
long to perform to be useful for either of these purposes. Thus, we
describe here our method for reducing the dimensionality to a set of
three templates which adequately describe the data.

Our procedure has two steps. First, we use principle component
analysis on the set of 225-coefficent vectors describing all the fits
in order to find a 10-dimensional subspace which still describes the
data. Second, within this 10-dimensional subspace we group the
galaxies into a set of ten groups, average the spectra within each
group, and pick the triple among these groups which best describes the
data. We adopted this procedure because we did not know of a
principled way in which to reduce the dimensionality without
sacrificing nonnegativity; however, it turns out that such methods do
exist, and we recommend those methods for future investigators ({\bf
refs from sam}) while retaining this method for the moment because it
appears to work for our purposes.

Because the galaxies all have very different bolometric fluxes, there
is a strong variation in the coefficient values simply due to the
total flux differing by large factors. Therefore, before performing
principal component analysis, we scale all the coefficients of each
galaxy by a single number (per galaxy) such that the galaxies all lie
exactly on a plane in coefficient space. We choose the plane to be the
plane on which the sum of the templates has a constant flux between
1500 \AA\ and 30000 \AA. After scaling the galaxies to this plane, we
perform principal component analysis to obtain the five most principal
axes, using the efficient method of \citet{roweis99a}. There is almost
no difference between the full fit shown above and that using the
first five eigencoefficients.

On the other hand, the five eigencomponents natural consist partly of
positive star-formation components and partly of negative
star-formation components. This means that we cannot simply sum them
nonnegatively to get a nonnegative star-formation history. In order to
get a set of components that consists purely of all-positive
star-formation histories, we consider the distribution of galaxies in
the five-component space. We perform a $K$-means grouping on this set
of points, which produces ten groups. This is simply a way of very
roughly mapping the locus on which the galaxies reside (it is
significantly nonlinear within the five-dimensional space, so we
cannot simply take fewer principal components). We then average the
spectra in each group. This gives ten spectra who roughly span the
space of galaxy types. Finally, we pick from among those ten the
triple which best describes the galaxies, in the sense of minimizing
$\chi^2$ in Equation \ref{chi2}.

Figure \ref{kfit3} shows the color residuals after fitting with the
final three templates.

[ new model: 3 PEGASE2 spectra; Figure show spectra, Figure stellar
  mass to light ratios ]

[ to deal with that for K-correction and photo-z, tweak ]

We recognize that this procedure is rather convoluted and a bit of a
hack. As noted above, we have adopted it simply as one way of
automatically getting a small set of templates that explain galaxy
broad-band fluxes as a sum of star-formation histories. There are
methods for directly performing principal component analysis subject
to the conditions that the components and the coefficients are all
positive, and we recommend the investigation of such methods to the
reader ({\bf ref}).

>>>>

\shortauthors{Blanton {\it et al.} (2000)}
\shorttitle{$K$-corrections and filter transformations}

@@

\title{$K$-corrections and filter transformations \\
from the restframe
ultraviolet to the near infrared}

\author{
Michael R. Blanton\altaffilmark{\ref{NYU}} and 
Samuel Roweis
%Tamas Budavari\altaffilmark{\ref{JHU}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%J.~Brinkmann\altaffilmark{\ref{APO}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Mamoru Doi\altaffilmark{\ref{Tokyo}},
%Daniel Eisenstein\altaffilmark{\ref{Arizona}},
%Masataka Fukugita\altaffilmark{\ref{CosmicRay},\ref{IAS}},
%James E. Gunn\altaffilmark{\ref{Princeton}},
%David W. Hogg\altaffilmark{\ref{NYU}}, and
%David J. Schlegel\altaffilmark{\ref{Princeton}}
%Julianne Dalcanton\altaffilmark{\ref{UW}},
%Jon Loveday\altaffilmark{\ref{Sussex}},
%Michael A. Strauss\altaffilmark{\ref{Princeton}},
%Mark SubbaRao\altaffilmark{\ref{Chicago}},
%David H. Weinberg\altaffilmark{\ref{Ohio}},
%John E. Anderson, Jr.\altaffilmark{\ref{Fermilab}},
%James Annis\altaffilmark{\ref{Fermilab}},
%Neta A. Bahcall\altaffilmark{\ref{Princeton}},
%Mariangela Bernardi\altaffilmark{\ref{Chicago}},
%Robert J. Brunner\altaffilmark{\ref{Caltech}},
%Scott Burles\altaffilmark{\ref{Fermilab}},
%Larry Carey\altaffilmark{\ref{UW}},
%Francisco J. Castander\altaffilmark{\ref{Chicago}, \ref{Pyrenees}},
%Andrew J. Connolly\altaffilmark{\ref{Pitt}},
%Istv\'an Csabai\altaffilmark{\ref{JHU}},
%Douglas Finkbeiner\altaffilmark{\ref{Berkeley}},
%Scott Friedman\altaffilmark{\ref{JHU}},
%Joshua A. Frieman\altaffilmark{\ref{Fermilab}},
%G. S. Hennessy\altaffilmark{\ref{USNO}},
%Robert B. Hindsley\altaffilmark{\ref{USNO}},
%Takashi Ichikawa\altaffilmark{\ref{Tokyo}},
%\v{Z}eljko Ivezi\'{c}\altaffilmark{\ref{Princeton}},
%Stephen Kent\altaffilmark{\ref{Fermilab}},
%G. R.~Knapp\altaffilmark{\ref{Princeton}},
%D. Q.~Lamb\altaffilmark{\ref{Chicago}},
%R. French Leger\altaffilmark{\ref{UW}},
%Daniel C. Long\altaffilmark{\ref{APO}},
%Robert H. Lupton\altaffilmark{\ref{Princeton}},
%Timothy A.~McKay\altaffilmark{\ref{Michigan}},
%Avery Meiksin\altaffilmark{\ref{Edinburgh}},
%Aronne Merelli\altaffilmark{\ref{Caltech}},
%Jeffrey A. Munn\altaffilmark{\ref{USNO}},
%Vijay Narayanan\altaffilmark{\ref{Princeton}},
%Matt Newcomb\altaffilmark{\ref{CarnegieMellon}},
%R. C. Nichol\altaffilmark{\ref{CarnegieMellon}},
%Sadanori Okamura\altaffilmark{\ref{Tokyo}},
%Russell Owen\altaffilmark{\ref{UW}},
%Jeffrey R.~Pier\altaffilmark{\ref{USNO}},
%Adrian Pope\altaffilmark{\ref{JHU}},
%Marc Postman\altaffilmark{\ref{STScI}},
%Thomas Quinn\altaffilmark{\ref{UW}},
%Constance M. Rockosi\altaffilmark{\ref{Chicago}},
%Donald P. Schneider\altaffilmark{\ref{PennState}}, 
%Kazuhiro Shimasaku\altaffilmark{\ref{Tokyo}},
%Walter A. Siegmund\altaffilmark{\ref{UW}},
%Stephen Smee\altaffilmark{\ref{Maryland}},
%Yehuda Snir\altaffilmark{\ref{CarnegieMellon}},
%Chris Stoughton\altaffilmark{\ref{Fermilab}},
%Christopher Stubbs\altaffilmark{\ref{UW}},
%Alexander S.~Szalay\altaffilmark{\ref{JHU}},
%Gyula P.~Szokoly\altaffilmark{\ref{Potsdam}},
%Aniruddha R.~Thakar\altaffilmark{\ref{JHU}},
%Christy Tremonti\altaffilmark{\ref{JHU}},
%Douglas L. Tucker\altaffilmark{\ref{Fermilab}},
%Alan Uomoto\altaffilmark{\ref{JHU}},
%Dan vanden Berk\altaffilmark{\ref{Fermilab}},
%Michael S. Vogeley\altaffilmark{\ref{Drexel}},
%Patrick Waddell\altaffilmark{\ref{UW}},
%Brian Yanny\altaffilmark{\ref{Fermilab}},
%Naoki Yasuda\altaffilmark{\ref{NAOJ}},
%and Donald G.~York\altaffilmark{\ref{Chicago}}
}

@@

\begin{abstract}
We present a method of developing model-based spectroscopic templates
to explain observed galaxy fluxes. Such templates are useful for
calculating $K$-corrections, calculating photometric redshifts, and
converting among observations of galaxies at various wavelengths. In
principle, we could apply this method to galaxy data at any redshift
and in any set of wavelengths. The implementation we present here is
suitable for ultraviolet, optical, and near-infrared
observations. Since we base the templates on stellar population
synthesis results, the results are intepretable in terms of
approximate stellar masses and star-formation histories. We present
templates fit with this method to data from GALEX, Sloan Digital Sky
Survey spectroscopy and photometry, the Two-Micron All Sky Survey, the
Deep Extragalactic Evolutionary Probe and Great Observatories Origins
Deep Survey.  In addition, we present software for using such data to
estimate $K$-corrections, photometric redshifts, evolution
corrections, stellar masses, and recent star-formation rates. This
software is suitable for analyzing many other data sets as well.
\end{abstract}code

@@

New surveys at a low and high redshift have provided us with estimates
of galaxy spectral energy distributions (SEDs) for an enormous number
of galaxies. When comparing populations of galaxies at different
redshifts in these surveys, we need to use comparable measurements of
the galaxy SEDs. However, different surveys use different bandpasses
and the restframe wavelengths of these bandpasses necessarily vary
with redshift. We need to be able to handle this heterogeneity in
order to make sensible comparisons among all of these new surveys.

In this paper, we present a method for doing so, by calculating
$K$-corrections between observed and desired bandpasses.  The
$K$-correction between a bandpass $R$ used to observe a galaxy at
redshift $z$ and the desired bandpass $Q$ is defined by the equation
(\cite{oke68a, hogg02a}):
\begin{equation}
\label{kcorrecteqn}
m_R = M_Q + \mathrm{DM}(z) + K_{QR}(z) - 5 \log_10 h 
\end{equation}
where $\mathrm{DM}(z) = 25 - 5\log_{10} (d_L / (h^{-1}{\mathrm{~Mpc}}))$ is the
bolometric distance modulus calculated from the luminosity distance
$d_L$, and $M_Q$ is the absolute magnitude. The absolute magnitude is
defined as the apparent magnitude an object would have if were
observed 10 pc away, in bandpass $Q$, at rest.  The traditional
definition of the $K$-correction takes $Q=R$. however, we note that in
practice many surveys do perform $K$-corrections from one observed
bandpass $R$ to another bandpass $Q$ in the rest frame. This practice
is particularly common when dealing with high redshift observations.
In addition to $K$-corrections, this method also provides an
interpretation of the data in terms of a physical model which
describes the stellar mass and star-formation history of each galaxy.

The templates are designed to work well for a wide range of data sets.
We constrain them with photometry and spectroscopy of the Sloan
Digital Sky Survey (SDSS; \citealt{york00a}) the Galaxy Evolution
Explorer (GALEX; {\bf ref})) in the ultraviolet, and the Two-Micron
All Sky Survey (2MASS; \citealt{strutskie97a}) in the near infrared
(NIR). In addition, at higher redshifts, we use constraints from the
Deep Extragalactic Evolutionary Probe 2 (DEEP2; {\bf ref} and the
Great Observatories Origins Deep Survey (GOODS; {\bf ref}).  These and
other data sets provide a huge set of information about galaxy colors
and spectra which we can use to help understand their star-formation
histories.

We note that the methods used here are generally useful. Using these
methods to develop templates has the advantage that one can easily
search a large space of models in order to find the few templates
necessary in order to explain a given set of data.

We also release the templates in electronic form as well as an
implementation of the methods used to fit the templates to data. This
software {\tt kcorrect v4\_1} is distributed on the World Wide
Web{\footnote {\tt http://cosmo.nyu.edu/blanton/kcorrect/}}. It
consists of a core C library which performs most of the complex and
computationally intensive tasks, plus an IDL library that provides a
high-level interface.  This software is an update of two major earlier
releases ({\tt v1\_16}, \citet{blanton03b}; {\tt v3\_2}). The
improvement over the latest version is twofold. First, we have
improved the templates such that they successfully fit galaxies in the
restframe UV, as observed by GALEX, DEEP2, and GOODS. Second, because
the templates are now completely model-based, the fits have a physical
interpretation in terms of a star-formation history. The IDL library
also has a number of new useful functions.

\section{Finding the templates}
\label{algorithm}

\subsection{Overview}

From the spectroscopic observations we know that galaxy spectra reside
in a low-dimensional locus. Principal Component Analysis (PCA) of
galaxy spectra, introduced by \citet{connolly94a} and applied many
times since then ({\bf more refs}; \citealt{yip05a}), demonstrates
that most of the variance in the distribution of galaxies in spectral
space can be explained using a few templates. This means that, in the
linear space of all possible spectra, galaxies exist in only a small
locus. Therefore, even with very heterogeneous data, we should be able
to determine the properties of this locus.

Here we present an approach to combining the heterogeneous data
described above in order to determine the properties of the locus of
galaxy spectra. Rather than taking the model-free approach used by
PCA, we here restrict the space of possible spectra to those predicted
from the high resolution stellar population synthesis model of
\citet{bruzual03a} and the emission line models of {\bf mappings
ref}.  This approach both constrains the problem appropriately and
yields a natural theoretical interpretation of the results in terms of
star-formation histories.

In a nutshell, our algorithm does the following. Given the
observations (and uncertainties of those observations) available for
each galaxy, it finds the nonnegative linear combination of $N$
template star-formation histories which best predict those
observations in the $\chi^2$ sense. Given the entire set of galaxy
observations available, it also fits for those template star-formation
histories. The technical name for this algorithm is Nonnegative Matrix
Factorization (NMF).

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, it has a natural physical
interpretation associated with that spectral subspace, which is the
corresponding subspace of all possible star-formation
histories. Second, it naturally handles data uncertainties and missing
data, which allows it to ignore that variation which is due purely to
statistical errors.  Third, it handles the complications of observing
galaxy spectra photometrically using broad-band filters.

In this section, we describe how to take the astrophysical
models and the astronomical observations, and to pose the problem in
the form needed for the purposes of NMF. In the Appendix we describe
how to solve to NMF problem.

\subsection{The models}

We begin with a basis set of 485 spectral templates. Of these, 450 are
a set of instantaneous bursts from \citet{bruzual03a}. There are 6
metallicities (0.005, 0.02, 0.2, 0.4, 1, and 2.5 times solar). For
each metallicity there are 25 ages (between 1 Myr and 13.75 Gyr,
spaced almost logarithmically in age).  For each age and metallicity
there are 3 choices of dust model: (1) no dust extinction; (2) $\tau_V
= 3$ dust with Milky Way type extinction (as input into the models of
\citealt{witt00a} with a ``homogeneous'' distribution and ``shell''
geometry); (3) $\tau_V = 3$ dust with SMC type extinction (from the
same models). We smooth each template to $300$ km s$^{-1}$ velocity
dispersion.

The remaining 35 templates are from MAPPINGS-III {\bf ref} models of
emission from ionized gas. We choose the predictions for an 8 Myr old
continuous star-formation history with 5 possible metalliciticites
(0.5, 0.2, 0.4, 1.0, and 2. times solar) and 7 possible ionization
parameters ($q= 5\times 10^6$, $10^7$, $2\times 10^7$, $4\times 10^7$,
$8 \times 10^7$, $1.5\times 10^8$, and $3\times 10^8$ cm s$^{-1}$). We
take the spectra (given as a set of delta functions) and smooth them
to $300$ km s$^{-1}$ velocity dispersion.

Let us refer to these 485 basis templates as $M_{\lambda,
j}(\lambda)$, expressed in units of ergs s$^{-1}$ \AA$^{-1}$.

We seek five templates $F_{\lambda, i}(\lambda)$ which are built from
nonnegative combinations of the original basis set of $N$ templates:
\begin{equation}
F_{\lambda, i}(\lambda) = \sum_j b_{ij} M_{\lambda, j}(\lambda),
\end{equation}
in units of ergs s$^{-1}$ \AA$^{-1}$.  For each galaxy $k$
we want our model ${\hat{F}_\lambda}(\lambda)$ for their spectrum to
be a nonegative sum of these five templates:
\begin{equation}
{\hat{F}}_{\lambda,_k}(\lambda) = \sum_i a_{ki} F_{\lambda,i}(\lambda).
\end{equation}

\subsection{The data}

The training set consists of:
\begin{enumerate}
\item SDSS spectroscopic data in the observed range $3800 < \lambda <
9000$ \AA, for 400 Luminous Red Galaxies between $0.15 < z < 0.5$ (LRGs;
\citealt{eisenstein02a}) and 1600 Main sample galaxies between $0.001
< z < 0.4$ (\citealt{strauss02a})
\item SDSS photometric data on an independent set of LRGs ($griz$
photometry only) and Main sample galaxies (using the full $ugriz$
photometry). We use 2000 LRGs and 7000 Main sample galaxies.  For
these galaxies we include 2MASS $JHK_s$ data ({\bf ref}) where
available.
\item GALEX DR1 far UV ($\sim 1500$ \AA) and near UV ($\sim 2300$ \AA)
photometry for objects with SDSS redshifts and $ugriz$ photometry (4000
galaxies; {\bf ref}).
\item The $BRI$ photometry for high redshift galaxies in the DEEP2 
DR1 release between $0.6 < z < 1.5$ (4000
galaxies; {\bf ref}).
\item The $BVizJHK_s$ photometry for GOODS galaxies between $0.5 < z < 2$ (1000
galaxies; {\bf ref}).
\end{enumerate}
The SDSS, 2MASS, and GALEX galaxies were selected from and the matches
were obtained by the New York University Value-Added Galaxy Catalog
(NYU-VAGC;
\citealt{blanton05b}).

The SDSS data processing consists of astrometry \citep{pier03a};
source identification, deblending and photometry \citep{lupton01a};
photometricity determination \citep{hogg01a}; calibration
\citep{fukugita96a,smith02a}; spectroscopic target selection
\citep{eisenstein01a,strauss02a,richards02a}; spectroscopic fiber
placement \citep{blanton03a}; and spectroscopic data reduction.  We
recalibrated our photometry using the ``ubercalibration'' procedure
described in \citet{blanton05b}.  Descriptions of these pipelines also
exist in \citet{stoughton02a}.  An automated pipeline called {\tt
idlspec2d} (in this case {\tt v4\_9}) measures the redshifts and
classifies the reduced spectra (Schlegel et al., in preparation).

Note that there are 20 different broadband photometric filters listed
above ($B$ in DEEP2 is different than the $B$ used by GOODS). The {\tt
kcorrect} product described in Section \ref{code} contains a
tabulation of the response functions for all of these filters.

\subsection{Comparing data and models}

The data consist of spectra and bandband photometric measurements of
galaxies at a number of redshifts, and we have to relate the models to
these measurements.

For the spectra, we take the observed spectrum
$f_{\lambda,k}(\lambda)$ (in ergs cm$^{-2}$ s$^{-1}$ \AA$^{-1}$) of
each galaxy $k$ at redshift $z$ (corresponding to a luminosity
distance $d_L(z)$ in cm; see \citealt{hogg99a}) and calculate the
restframe luminosity per unit wavelength:
\begin{equation}
{{F}}_{\lambda, k}(\lambda) = 
{{f}}_{\lambda, k}[\lambda (1+z)](1+z) (4 \pi d_L^2),
\end{equation}
That is, the spectrum is shifted due to the redshift, while the
integral of the numerator over wavelength (the total luminosity) is
constant with redshift, and the total flux is related to the total
luminosity by the inverse square law.  In addition, we smooth each
spectrum by such an amount that, given its estimated velocity
dispersion, its total velocity dispserion after smoothing is 300 km
s$^{-1}$.

Note that if we discretize the spectra to wavelengths $\lambda_l$, the
relationship between the predicted SED for a galaxy $k$ and the basis
set $M_{jl}$, becomes simply:
\begin{equation}
\hat{F}_{kl} = \sum_{ij} a_{ki} b_{ij} M_{jl}
\end{equation}

If one has a spectrum for a galaxy the
expression for the contribution to $\chi^2$ from each wavelength
$\lambda_l$ in the spectrum is then quite simple:
\begin{eqnarray}
\label{chi2spec}
\chi_{kl}^2 &=& \left[\frac{F_{\lambda,k}(\lambda_l) -
{\hat{F}}_{\lambda,k}(\lambda_l)} 
{\sigma^2_k(\lambda_l)}\right]^2 \cr
&=&
\left[\frac{F_{\lambda,k}(\lambda_l) -
\sum_{ij} a_{ki} b_{ij} M_{jl}
}{\sigma^2_k(\lambda_l)}\right]^2. 
\end{eqnarray}

Comparing observed broadband flux measurements is a bit more
complicated, because it is a projection of the spectrum onto the
broadband filter $p$ at the observed redshift $z$. Instead of
adjusting the observed fluxes as we could so easily do for the spectra
above, for the photometry we express the models in terms of predicted
broadband fluxes.

Here we express the flux for each galaxy $k$ in units of AB maggies
$\mu_p$, which are defined as what one would measure in bandpass $p$
relative to the AB standard source. For example, if we transform our
spectral energy density basis function $M_{\lambda, j}(\lambda)$ to a flux
density $m_{\lambda, j}(\lambda)$, we can calculate the contribution of
that basis function to the predicted maggies as:
\begin{equation}
\label{maggies}
\eta_{jp} = 
\frac{\int_0^{\infty} d\lambda \lambda R_p(\lambda) {m}_{\lambda, j}(\lambda)}
{\int_0^\infty d\lambda \lambda R_p(\lambda) f_{\lambda,
\mathrm{AB}}(\lambda) }
\end{equation}
Here, the response function $R_p(\lambda)$ is proportional to the
contribution to the detector signal of a photon with wavelength
$\lambda$ entering the Earth's atmosphere (or entering the telescope
for a space telescope). The AB standard source is $f_{\lambda,
\mathrm{AB}}(\lambda) d\lambda = f_{\nu,
\mathrm{AB}} (\nu) d\nu$ and $f_{\nu, \mathrm{AB}}=3631$ Jy $= 3.631 \times
10^{-20}$ erg s$^{-1}$ cm$^{-2}$ Hz$^{-1}$. Of course maggies $\mu$
are related to magnitudes $m$ as:
\begin{equation}
m = -2.5 \log_{10} \mu, 
\end{equation}
such that the AB standard source would (if it existed) have $\mu=1$
and $m=0$ for all bandpasses.

In this context it is worth noting that many authors
(e.g. \citealt{bessell90a}) tabulate the contribution to the detector
signal per {\it unit of energy} in photons of wavelength $\lambda$
instead of per {\it photon} with wavelength $\lambda$.  We will refer here
to that quantity as $R_p'(\lambda)$, though in the literature it is
often referred to without a prime (and without any explicit
definition!). Clearly $R_p(\lambda) \propto R_p'(\lambda)/\lambda$,
since the higher the frequency of the photon, at a fixed response per
unit energy there is a higher response per photon. With this
substitution, one can reexpress Equation
\ref{maggies} appropriately in terms of $R_p'(\lambda)$.  
Generally, though not universally, authors tabulate $R_p'(\lambda)$
for bandpasses whose standards were originally calibrated using
energy-counting devices rather than photon-counting devices. However,
from the point of view of the analysis of the observations it is
irrelevant what the devices used for the standards and the
observations are, as long as one calculates $R_p(\lambda)$ and uses
Equation \ref{maggies}.

The prediction for the broadband fluxes from Equation
\ref{maggies} is only for a specific redshift $z$. It turns out to
simplify our mathematics to calculate the projection of each basis
function $j$ onto each filter $p$ for a grid of redshifts (in this
case spaced by 0.005 between redshifts 0 and 2). Thus, below we will
take $p$ to index all the filters at all such redshifts.

Just as before, we can now write down the relationship between the
predicted broadband flux and the basis set $\mu_{jp}$, in the bandpass
and redshift corresponding to the index $p$, for a galaxy $k$:
\begin{equation}
\hat{\mu}_{kp} = \sum_{ij} a_{ki} b_{ij} \eta_{jp}
\end{equation}
